PAGAI : a path sensitive static analyser

### 2012-05-23 papers/TAPAS2012/pagai.tex

\lstinline|assert| and \lstinline|assume| are implemented as C macros. \lstinline|assert(x)| is roughly defined as \lstinline|if (! x) __assert_fail();|, and the tool just tests for the reachability of \lstinline|__assert_fail();|: if it is unreachable, then the assertion is true.

@@

Let us briefly comment the results given in more details in Table~\ref{tab:techniques}.
\emph{Guided Static Analysis} from \citet{DBLP:conf/sas/GopanR07} improves the
result of the classical Abstract Interpretation in $2.72\%$ of the control points
in $P_R$.
\emph{Path-focusing} from \citet{Monniaux_Gonnord_SAS11} gives
statistically better results, and finds better invariants than \emph{Guided
Static Analysis} in $9.89\%$ of the cases. However, it also loses precision in an
important number ($2.68\%$) of control points.
Finally, our combined technique gives the most promising results, since it is
statistically more precise than the other techniques.

The analysis using disjunctive invariants greatly improves the
precision of the analysis (for $12.06\%$ of the control points in $P_R$), at an
acceptable time cost (see Table \ref{tab:time}).

@@

Table \ref{tab:time} gives the execution time of the different analysis
techniques. It is interesting to see that
\emph{Path-focusing} is sometimes faster than the classical
algorithm.
This seems due to the fact that these algorithms start on a small subset
of the graph, and then may need fewer iterations on the complete graph.
The techniques we introduce in this article are slower, but speed is still on the same order of magnitude as the older ones.

@@

\subsection{Precision of Abstract Domains}
\label{sec:compare_domains}
\begin{table}[!tb]
	\tiny
\begin{center}
\setlength{\tabcolsep}{0.75ex}
\begin{tabular}{|l
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}|} \hline
\multicolumn{1}{|c|}{\textbf{Benchmark}}
& \multicolumn{3}{c|}{\textbf{PK/OCT}}
& \multicolumn{3}{c|}{\textbf{PK/BOX}}
& \multicolumn{3}{c|}{\textbf{OCT/BOX}}
& \multicolumn{3}{c|}{\textbf{PK/PKEQ}}
& \multicolumn{3}{c|}{\textbf{PK/GRID}}
& \multicolumn{3}{c|}{\textbf{POLY*/POLY}} \\ %\cline{2-16}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.} \\
 \hline
 \input{table_domain.tex}
\end{tabular}
\end{center}
\caption{Results of the comparison of the various abstract domains, when using
the same technique (G+PF). We used as abstract domains Convex Polyhedra (PK and
POLY),
Octagons (OCT), intervals (BOX), linear equalities (PKEQ) and linear congruences
(GRID). The last column compares the domain of Convex Polyhedra with the
improved widening operator from \citet{BagnaraHRZ05SCP} (POLY*), and Convex Polyhedra
using the classical widening operator (POLY). POLY and POLY* use the Parma
Polyhedral Library \cite{BagnaraHZ08SCP}.
$\subsetneq$, $\supsetneq$ and ``unc.'' are defined as in Tab.~\ref{tab:techniques}.}
\label{tab:domain}
\end{table}

@@

The Lapack/Blas benchmarks are unusual compared to the other programs. These libraries perform matrix computations, using nested loops over indices; such programs are the prime target for polyhedral loop optimization techniques and it is therefore unsurprising that polyhedra perform very well over them.

The analysis of linear equalities (PKEQ) performs very fast compared
to other abstract domains, but yields very imprecise invariants:
it only detects relations of the form $\sum_i a_i x_i = C$ where $a_i$ and $C$ are constants.
The comparison between the domain of convex polyhedra and linear congruences
shows the percentage of points where the use of congruences refines the
invariant. This is the case for $1.04\%$ of the control points in $P_R$.

Finally, we evaluated the benefits of the improved version of the widening
operator for convex polyhedra from \citet{BagnaraHRZ05SCP}, compared to the
classical widening. We found that the improved version from
\citet{BagnaraHRZ05SCP} yields more precise invariants for $6.58\%$ of the
control points in $P_R$.

It is not totally relevant to compare by inclusion the abstract values obtained
by the various analysis techniques. Indeed, a slightly smaller invariant may not
always be useful to prove the desired properties.
A future work is thus to experiment with
better comparison metrics, such as:
\begin{itemize}
	\item the number of \emph{assert} that have been proved in the code.
		Unfortunately, it is difficult to find good benchmarks or real life
		programs with many \emph{assert} statements.
	\item the number of false alarms in a client analysis that detects array
		bound violations, arithmetic overflows, etc.
\end{itemize}

>>>>

\lstinline|assert| and \lstinline|assume| are implemented as C macros.
\lstinline|assert(x)| is roughly defined as \lstinline|if (! x) __assert_fail();|, and the tool just tests for the reachability of \lstinline|__assert_fail();|: if it is unreachable, then the assertion is true.
\lstinline|assume| works with the same principle, and is defined as \lstinline|if (! x) __assumption_declared()|. Since \lstinline|__assert_fail| and
\lstinline|__assumption_declared| are \emph{noreturn} functions, the code after
the test verifies \lstinline|x|.

@@

Let us briefly comment the results given in more details in Table~\ref{tab:techniques}.
\emph{Guided Static Analysis} from \citet{DBLP:conf/sas/GopanR07} improves the
result of the classical Abstract Interpretation in $2.21\%$ of the control points
in $P_R$.
\emph{Path-focusing} from \citet{Monniaux_Gonnord_SAS11} 
finds better invariants in $4.13\%$ of the control points.

However, these two techniques also lose precision in an important number
($4.64\%$ for G, $5.14\%$ for PF) of control points, and obtain worse results
than the classical many times. This result is unexpected, and could be partially
explained by bad behaviour of the widening operator.

Finally, our combined technique gives the most promising results, since it is
statistically more precise than the other techniques. It improves the precision
of the inductive invariant in $8.29\%$ to $9.86\%$ of the control points
compared to the three previous techniques. Still, we obtain worse result in a
non-negligible number of cases ($2.02\%$).

The analysis using disjunctive invariants greatly improves the
precision of the analysis (for $14.46\%$ of the control points in $P_R$ compared
to G+PF), at the expense of a higher time cost (see Table \ref{tab:time}).
It also gives worse results in $6.85\%$ of the points, most probably because of
a non optimal choice of the $\sigma$ function, detailed in
\cite{Henry_Monniaux_Moy_SAS12}.

@@

Table \ref{tab:time} gives the execution time of the different analysis
techniques. It is interesting to see that
\emph{Path-focusing} is sometimes faster than the classical
algorithm.
This seems due to the fact that this algorithm
computes inductive invariant on a small number of control points compared to
classical approaches, thus leading to fewer operations over abstract values.

@@

\subsection{Precision of Abstract Domains}
\label{sec:compare_domains}
\begin{table}[!tb]
	\tiny
\begin{center}
\setlength{\tabcolsep}{0.75ex}
\begin{tabular}{|l
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}%
|D{.}{.}{2}D{.}{.}{2}D{.}{.}{2}|} \hline
\multicolumn{1}{|c|}{\textbf{Benchmark}}
& \multicolumn{3}{c|}{\textbf{PK/OCT}}
& \multicolumn{3}{c|}{\textbf{PK/BOX}}
& \multicolumn{3}{c|}{\textbf{OCT/BOX}}
& \multicolumn{3}{c|}{\textbf{PK/PKEQ}}
& \multicolumn{3}{c|}{\textbf{PK/PKGRID}}
& \multicolumn{3}{c|}{\textbf{POLY/POLY*}} \\ %\cline{2-16}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.}
& \multicolumn{1}{c}{$\subsetneq$} & \multicolumn{1}{c}{$\supsetneq$} & \multicolumn{1}{c|}{unc.} \\
 \hline
 \input{table_domain.tex}
\end{tabular}
\end{center}
\caption{Results of the comparison of the various abstract domains, when using
the same technique (G+PF). We used as abstract domains Convex Polyhedra (PK and
POLY),
Octagons (OCT), intervals (BOX), linear equalities (PKEQ) and the reduced
product of NewPolka convex polyhedra with linear congruences
from the Parma Polyhedra Library \cite{BagnaraHZ08SCP}.
(PKGRID). The last column compares the domain of Convex Polyhedra with the
improved widening operator from \citet{BagnaraHRZ05SCP} (POLY*), and Convex Polyhedra
using the classical widening operator (POLY). POLY and POLY* use the PPL\cite{BagnaraHZ08SCP}.
$\subsetneq$, $\supsetneq$ and ``unc.'' are defined as in Tab.~\ref{tab:techniques}.}
\label{tab:domain}
\end{table}

@@

The Lapack/Blas benchmarks are unusual compared to the other programs. These
libraries perform matrix computations, using nested loops over indices; such
programs are the prime target for polyhedral loop optimization techniques and it
is therefore unsurprising that polyhedra and octagons perform very well over them.

The analysis of linear equalities (PKEQ) performs very fast compared
to other abstract domains, but yields very imprecise invariants:
it only detects relations of the form $\sum_i a_i x_i = C$ where $a_i$ and $C$ are constants.

Using the reduced product of convex prolyhedra with linear congruences
(PKGRID) improves the analysis by $2.52\%$.

Finally, we evaluated the benefits of the improved version of the widening
operator for convex polyhedra from \citet{BagnaraHRZ05SCP}, compared to the
classical widening. We found that the improved version from
\citet{BagnaraHRZ05SCP} yields more precise invariants for $3.70\%$ of the
control points in $P_R$.

\subsection{Future Work}
It is not totally relevant to compare by inclusion the abstract values obtained
by the various analysis techniques. Indeed, a slightly smaller invariant may not
always be useful to prove the desired properties.
A future work is thus to experiment with
better comparison metrics, such as:
\begin{itemize}
	\item the number of \emph{assert} that have been proved in the code.
		Unfortunately, it is difficult to find good benchmarks or real life
		programs with many \emph{assert} statements.
	\item the number of false alarms in a client analysis that detects array
		bound violations, arithmetic overflows, etc.
\end{itemize}
